---
title: "Animating Plots in R"
author: "Stephanie Kirmer"
date: "June 2019"
output: 
    ioslides_presentation:
        widescreen: true
---


# Agenda

**Animation: When and Why?**

**Introducing the Tools**

**Hands-On Tutorial**


## So You're Thinking About Animating

What are some reasons to animate a plot?

### **Bad Reasons**

- It's neat  
- I want to show off  

### **Good Reasons**

- I have four or more data dimensions   
- I want to display many levels of a single dimension  


## Examples of Good Cases

- data changing over a long period of time
- groups of datapoints varying systematically

Also: it is not necessary to see all the data points at the same time

### Ask Yourself
- What information do you want the audience to take away?  

# The Tools


## Packages to Use
```{r}
library(ggplot2) # make plots
library(gganimate) # animate the plots
library(gifski) # render gifs
library(transformr) # insert smooth transformations
```

## Key Concepts

### **Frame**
- Animations are a series of single images (frames) strung together

### **Transition**
- How the visual rendering moves from frame to frame

### **State**
- The current frame at any given point in time

### **Enter/Exit**
- How datapoints with no predecessor or follower are handled



# Tutorial

## Project Overview

Using data on displaced persons living in South Africa, show patterns and trends in refugee movement over twenty-plus years.

**Task:** show complex data across many levels of a facet  
**Problems:** Facet grid and facet wrap rapidly get out of their depth  

Data Source: UNHCR via data.world: [https://data.world/unhcr/244d728f-06e6-4c25-8eed-69e8fe3bffe0](UNHCR's populations of concern residing in South Africa)

## Load Data

Grab dataset of displaced persons living in South Africa
```{r, echo = FALSE}
dtset <- read.csv("data/unhcr-time-series-residing-zaf-csv-1.csv", skip = 1)
names(dtset) <- gsub("\\.", "_", names(dtset))

knitr::kable(head(dtset, 4))
```


## Mutate Data

Select the top 10 countries of origin for each year, apply some filters.

```{r, message=FALSE, error=FALSE, warning=FALSE}
library(magrittr)
library(dplyr)

plotDT <- dtset %>%
  filter(X_country_origin != "Various/Unknown") %>%
  filter(X_population_type == "Refugees (incl. refugee-like situations)") %>%
  filter(X_date_year > 1995) %>%
  group_by(X_date_year) %>%
  mutate(rank = rank(-X_affected, ties.method = "first") * 1) %>%
  ungroup() %>%
  filter(rank <= 10) %>%
  data.frame()
```

```{r, echo = FALSE}
plotDT$X_country_origin <- ifelse(plotDT$X_country_origin == "Serbia and Kosovo (S/RES/1244 (1999))", "Serbia/Kosovo", as.character(plotDT$X_country_origin))
plotDT$state_time <- ifelse(plotDT$X_date_year == max(plotDT$X_date_year), 5, 1)
```

## Our New Data

```{r, echo=FALSE}
knitr::kable(head(plotDT, 4))
```

## Grouped Bar Plot - Code

```{r}
baseplot1 <- ggplot(plotDT, 
aes(X_date_year, 
    group = X_country_origin, 
    fill = as.factor(X_country_origin), 
    color = as.factor(X_country_origin)))+
theme_bw()+
theme(legend.position = "bottom")+
geom_bar(aes(y = X_affected), stat = "identity", position = "dodge")

```

## Grouped Bar Plot - Rendered

Ew.

```{r, echo=FALSE, fig.width = 10}
baseplot1
```


## Adjustment 1: Flip Axis

```{r, fig.width = 8, fig.height=4}
baseplot2 <- baseplot1 +
    coord_flip(clip = "off", expand = FALSE, ylim = c(0, 50000))


```

## Render

```{r, echo=FALSE, fig.width = 10}
baseplot2
```


## Adjustment 2: Add Descriptive Labels

```{r, fig.width = 8, fig.height=3}
baseplot2 <- baseplot2 +
geom_text(aes(y = 0, label = paste(X_country_origin, " ")), vjust = 0.2, hjust = 1) +
geom_text(aes(y = X_affected, 
    label = as.character(X_affected)), 
    color = "black", vjust = 0.2, hjust = -.5)

```

## Render

```{r, echo=FALSE, fig.width = 10}
baseplot2
```


## Adjustment 3: Stop the bar grouping (dodge), switch axis to country from year

Can't use the same stub now.
```{r}
baseplot3 <- ggplot(plotDT, 
aes(x=rank, 
    group = X_country_origin, 
    fill = as.factor(X_country_origin), 
    color = as.factor(X_country_origin)))+
theme_bw()+
theme(legend.position = "bottom")+
geom_text(aes(y = 0, label = paste(X_country_origin, " ")), vjust = 0.2, hjust = 1) +
geom_text(aes(y = X_affected, 
    label = as.character(X_affected)), 
    color = "black", vjust = 0.2, hjust = -.5)+
coord_flip(clip = "off", expand = FALSE, ylim = c(0, 50000)) +
geom_bar(aes(y = X_affected), stat = "identity", position = "identity")

```

## Render

```{r, echo=FALSE, fig.width = 10}
baseplot3
```


## Check in: baseplot2 vs baseplot3

Left side: grouped bar, sorted by year  
Right side: not grouped bar, sorted by rank

```{r, echo = FALSE, fig.width = 10}
library(patchwork)
baseplot2 + baseplot3 + plot_layout(ncol = 2)
```


## Adjustment 4: Adjust margins, fix axis text, drop legend

If your stub has a `theme()` segment, applying a new one will overrule it.
```{r}
baseplot3 <- baseplot3 + 
theme(legend.position = "none",
    axis.ticks.y = element_blank(),  # These relate to the axes post-flip
    axis.text.y  = element_blank(),  # These relate to the axes post-flip
    axis.title.y = element_blank(),
    plot.margin = margin(1,1,1,5, "cm"))

```

## Render

```{r, echo=FALSE, fig.width = 10}
baseplot3
```


## Adjustment 5: Prettify Y, reverse X direction

```{r}

baseplot3 <- baseplot3 +
scale_y_continuous(labels = scales::comma) +
scale_x_reverse()

```


## Render

Now we have the different "frames" all layered on top of each other.
```{r, echo = FALSE, fig.width = 10}
baseplot3
```

## Decision Point: How to Proceed?
We need to present the data on an annual basis in a way that tells us:

1. What locations refugees come from  
    - These may be places with high instability or political turmoil at a given time

2. How the top locations within a year compare to the other top 10 locations

## Faceting?

```{r, echo = FALSE, fig.width = 10}
baseplot3+facet_wrap(X_date_year~.)
```

No.


## Faceting?

```{r, echo = FALSE, fig.width = 10}
baseplot3+facet_grid(X_date_year~.)
```

Definitely not.

# Develop the Animation

## Basic Version


```{r}

animp <- baseplot3 +
transition_states(
    X_date_year)

```

## Render

- Moving between frames without transition smoothing   
- Audience can't tell what the frames represent  
- Y axis needs labeling  

```{r, fig.width = 10, fig.height=6, echo = FALSE}

animate(animp, fps = 10, duration = 20) 

```


## Second Iteration

Added a descriptive title/label that indicate the year of the frame

```{r}

animp <- baseplot3 +
  labs(title = "Refugees Residing in South Africa by Origin, {closest_state}"
    , y="Affected Persons")+
  transition_states(X_date_year)

```

## Render

```{r, fig.width = 10, fig.height=6, echo = FALSE}

animate(animp, fps = 10, duration = 20) 

```

## Transition Options: Exit/Enter

Shrink and grow on exit and enter
```{r}

animp <- baseplot3 +
labs(title = "Refugees Residing in South Africa by Origin, {closest_state}"
  , y="Affected Persons")+
transition_states(X_date_year)+
enter_grow() +
exit_shrink()

```


## Render

Interesting, but probably not serving the project objectives

```{r, fig.width = 10, fig.height=6, echo = FALSE}

animate(animp, fps = 10, duration = 20) 

```


## Transition Options: Easing

Ease between positions (moving on page, not exiting or entering)
```{r}

animp <- baseplot3 +
labs(title = "Refugees Residing in South Africa by Origin, {closest_state}"
  , y="Affected Persons")+
transition_states(X_date_year)+
ease_aes('quartic-in-out')

```


## Render

Hopefully, the movement feels less abrasive to the eye

```{r, fig.width = 10, fig.height=6, echo = FALSE}

animate(animp, fps = 10, duration = 20) 

```

## Transition Options: Easing
For fun, let's try "back" to see a springier approach

```{r}

animp <- baseplot3 +
labs(title = "Refugees Residing in South Africa by Origin, {closest_state}"
  , y="Affected Persons")+
transition_states(X_date_year)+
ease_aes('back-in-out')

```


## Render

Feels a little cartoony- interesting, but again perhaps not what we need

```{r, fig.width = 10, fig.height=6, echo = FALSE}

animate(animp, fps = 10, duration = 20) 

```

## Transition Options: Timing

- Set pace for the states/transitions  
- Ensure no smoothing between the end and restarting  

```{r}

animp <- baseplot3 +
labs(title = "Refugees Residing in South Africa by Origin, {closest_state}"
  , y="Affected Persons")+
transition_states(
    X_date_year,
    transition_length = 4, 
    state_length = c(rep(3, 21), 30), 
    wrap = FALSE)+
ease_aes('quartic-in-out')+
enter_fade() +
exit_fade()

```


## Render

Slower pace feels smoother

```{r, fig.width = 10, fig.height=6, echo = FALSE}

animate(animp, fps = 10, duration = 20) 

```

## Dissect Animation Code: Transition States

```{r, eval=FALSE}
transition_states(
    X_date_year,
    transition_length = 4, 
    state_length = c(rep(1, 21), 30)
  , wrap = FALSE)
```

Assign the frame unit- here we use the year.  

Transition length is the period of time we use for change from frame to frame  
- slow it down if you want a smooth looking animation   

State length is the period of time where the frame stays static.  
- here I am making the very last frame stay static longer   

Wrap determines whether to apply transition smoothing between the end and restarting  

Experiment with these settings to get the look you want!  


## Render the Final Animation

```{r, fig.width = 10, fig.height=6, eval=TRUE} 
animate(animp, fps = 10, duration = 20, width = 700, height = 400) 
```


## Save the Animation

```{r, eval=FALSE} 
anim_save(filename = "final_race_plot.gif")
```

## Lessons to Take Away

Figure out the in/out transitions that make sense and are not misleading

Give the eye enough time to understand each state

## References

https://gganimate.com/

https://ggplot2.tidyverse.org/

